Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-09 programlogic.tex
finish pass on section 4, including tentative text on the control transfer axioms

 Definition L4_L1_PointsTo (maddr l4e l3e l2e l1e paddr :word 64) : vProp $\Sigma$ := $\lambda$ cr3val.
  $\ulcorner$ entry_present l4e $\land$ entry_present l3e $\land$ entry_present l2e $\land$ entry_present l1e$\urcorner$ $\ast$
  (l4M52 maddr cr3val) $\sim$ (l4off maddr cr3val) $\mapsto_{a}$ {q1}  l4e  $\ast$
  (l3M52 maddr l4e) $\sim$ (l3off maddr l4e)  $\mapsto_{a}$ {q2}  l3e $\ast$ 
  (l2M52 maddr l3e) $\sim$ (l2off maddr l3e) $\mapsto_{a}$ {q3}  l2e $\ast$
  (l1M52 maddr l2e) $\sim$ (l1off maddr l2e) $\mapsto_{a}$ {q4}  l1e.
\end{lstlisting}
\vspace{-1em}
\caption{A Strong Virtual Points-to Relation}
  \label{fig:strongvirtualpointsto}
\end{figure*}

\subsection{An Overly-Restrictive Definition for Virtual Memory Addressing}
\label{sec:overly-restrictive}
A natural definition for a virtual points-to that depends on the current address space is to
require that in order for a virtual address \textsf{va} to point to a value \textsf{v}, the assertion contains
partial ownership of the physical memory involved in the page table walk that would translate \textsf{va} to
its backing physical location --- with locations existentially quantified since a virtual points-to should not assert
\emph{which} locations are accessed in a page table walk.
Figure \ref{fig:strongvirtualpointsto} gives such a definition.
It asserts the existence of four page table entries, one at each translation level, and via \lstinline|L4_L1_PointsTo|
asserts from the current \lstinline|cr3| value, taking the offsets from \textsf{va} (as in Figure \ref{fig:pagetables})
to index each level, finding the existentially quantified entries in each level, 
ending with the L1 entry, whose physical page base is added to the page offset of \textsf{va} to obtain the physical address
 --- and there (\lstinline|addr_L1(va,l1e)|) in physical memory, is the value.
Most of the definition lives directly in \textsf{vProp}, using the separation logic structure lifted from Iris's \textsf{iProp}.
\todo[inline]{Colin notes: the \lstinline|addr_L1| mentioned in the paragraph above doesn't appear in Figure 2. paddr isn't mentioned
in the definition of \lstinline|L4_L1_PointsTo| in the figure. Shouldn't there be a pure \lstinline|addr_L1(va,l1e)=paddr| in there?
}

@@

  \begin{mathpar}
  \inferrule{ }{\mathsf{GhostMap}(\gamma,\theta) \ast \ghostmaptoken{\gamma}{\mathsf{pa}}{\mathsf{va}} \vs 
  \mathsf{GhostMap}(\gamma,\theta[\mathsf{pa}\mapsto\mathsf{va'}]) \ast \ghostmaptoken{\gamma}{\mathsf{pa}}{\mathsf{va}'} 
  }
  \and
  \inferrule{ }{\mathsf{GhostMap}(\gamma,\theta) \ast \ghostmaptoken{\gamma}{\mathsf{pa}}{\mathsf{va}} \wand \ulcorner \theta(\mathsf{pa})=\mathsf{Some}(\matsf{va}) \urcorner }
  \end{mathpar}

  % Updating the virtual-to-physical (ghost) mapping for a virtual address $va$
  % requires full ownership of the token. 
  Then the \emph{virtual memory manager's invariant} ensures that for each $\ghostmaptoken{\gamma}{va}{pa}$ mapping
  in this map, there are \emph{physical} resources sufficient to ensure that address translation for $va$
  will resolve in the hardware to $pa$ via --- $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$.
  This allows modification of the page tables themselves as long as the changes do not invalidate overall virtual-to-physical
  translation.
  Consistency follows from replacing the page table walk with ownership of the ghost map token for the relevant ghost map
  entry, preventing the overall translation from being changed separately from the virtual points-to.

@@

  \begin{figure*}
  \footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}(\ptablestore,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}(\ptablestore,m) \ast \\
 \qquad\qquad\qquad \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e l3e l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup} (\ptablestore,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
    %\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}
    \ptableabswalk{\delta,\theta}

\end{array}
\]
\caption{Global Address-Space Invariant with a fixed global map of address-space names $m$}
  \label{fig:peraspaceinvariant}
  \end{figure*}

@@

% In Figure \ref{fig:addrswitch}, we see two different address-spaces rooted at R0 and R1 in the pre/post condition. To be put into words, while switching an address-space by executing \lstinline|mov_ctl|) updating the  \lstinline|cr3|, it is expected to mention resources of two different address-spaces in the specification. In the precondition,  \lstinline|cr3| is loaded with R0 and all the virtual-pointsto assertions $a_0 \mapsto_{\textsf{v}}p_0$ and $a_1 \mapsto_{\textsf{v}}p_1$ are in the current-view of the memory and valid in the ambient logic \textsf{vProp}. The other address-space with the root address R1 has its own resources ($b_0 \mapsto_{\textsf{v}}p_3$ and $b_1 \mapsto_{\textsf{v}}p_4$) which are invalid under the current address space rooted at R0. This pattern of specification requires resources being mentioned under the \textit{world} they are valid, i.e. a modal-framing within the specification. Once the current address-space becomes the one with root address R1, i.e.  \lstinline|cr3| is loaded with R1, the context-resources $b_0 \mapsto_{\textsf{v}}p_3$ and $b_1 \mapsto_{\textsf{v}}p_4$ are loaded into the current-view of the memory, and all the other resources that are valid under R0 are introduced into the modal context rooted at R0
\begin{figure}
\footnotesize
\[\begin{array}{c}
    [r](\mathsf{P}) \;: \mathsf{vProp} \; \Sigma \; \stackrel{\triangle}{=} \; \lambda \_,\; \mathsf{P}~\mathsf{r} \vspace{0.5em}\qquad
    \textsf{Fact P} \stackrel{\triangle}{=} \;\forall \; ,r \; r' \ldotp  \textsf{P r} \dashv\vdash \textsf{P r'}\\
    \inferrule{}{(\textsf{P} \vdash \textsf{Q}) \vdash  ([r](\textsf{P}) \vdash  [r](\textsf{Q}))}\qquad
    \inferrule{}{[r](\textsf{P} \ast \textsf{Q}) \dashv\vdash ([r](\textsf{P}) \ast [r](\textsf{Q})}\vspace{0.5em}\qquad
    \inferrule{}{[r](\textsf{P}\land \textsf{Q}) \dashv\vdash [r](\textsf{P})\land [r](\textsf{Q})}\\
    \inferrule{}{[r](\textsf{P}\lor \textsf{Q}) \dashv\vdash [r](\textsf{P})\lor [r](\textsf{Q})}\qquad
    \inferrule{}{\textsf{Fact P} \vdash  [r](\textsf{P}) \dashv\vdash (\textsf{P})} %(\textsf{vProp }\Sigma))\\
\end{array}\]
\vspace{-1em}
  \caption{Other-space Modality and Its Laws}
  \label{fig:modaldef}
  \end{figure}
Figure \ref{fig:modaldef} gives the definition of our modal operator for asserting truth of a modal
(address-space-contingent) assertion \emph{in another address space}, which we call
the \emph{other-space} modality. The definition itself is not
particularly surprising --- as our modal assertions are semantically predicates on a page table root (physical)
address, the assertion $[r](P)$ is a modal assertion which ignores the (implicit) current page table root,
and evaluates the truth of $P$ as if $r$ were the page table root. This simple definition is 
what we need in order for assertions in one address space to talk about assertions in another.
\looseness=-1

@@

We can prove that this modality follows certain useful laws: that its truth is independent of the address
space in which it is considered,  it distributes over various logical connectives, and it follows the rule of
consequence.
We call \textsf{vProp} assertions whose truth is independent of the current address space
\textsf{Fact}s; these include other-space assertions, physical memory points-tos, and register assertions.
\textsf{Fact}s can move in and out of other-space modalities freely.

@@

\subsection{The Impact of Changing Address Spaces on Modal Assertions}
% \subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
% \subsubsection{Issues with Using Hoare Triples in Address-Space Specification}
\label{sec:issues}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|cr3|. Consider the hypothetical rule
\begin{mathpar}
\inferrule*[left=Broken]{ }{
  \{P \ast \textsf{cr3}\mapsto_{r} \;r_1 \ast r \mapsto_{r} \;r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast \textsf{cr3} \mapsto_{r} r_2 \ast r \mapsto_{r} r_2 \ast Q\}
}
\end{mathpar}
which is slightly simplified (by dropping per-address-space resources) for clarity in making the following point.
This rule captures the intuitive change of address space in a Hoare triple.
This interacts quite poorly with the traditional frame
rule and the modal flavor of virtual points-to assertions.
If we abbreviate the pre- and post-conditions above as \textsf{Pre} and \textsf{Post}:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\textsf{Pre}\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{\textsf{Post}\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \textsf{Pre}\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast \textsf{Post}\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. 
The essential problem is that the frame rule is motivated by local reasoning about local updates, but
a switch of address space is a \emph{global} change, since it may invalidate information about virtual addresses.
Thus framing around \lstinline|cr3| updates is unsound, which has important impacts on our formal development.

@@

This is then how we justify our actual rule for \lstinline|cr3| updates, which is \emph{almost}
\begin{mathpar}
\inferrule*[left=ChangeAddressSpace]{
  \{[r_1](P) \ast cr3 \mapsto_{r} r_2 \ast r \mapsto_{r} r_2 \ast Q\}\;\overline{is}
}{
  \{P \ast cr3 \mapsto_{r} r_1 \ast r \mapsto_{r} r_2 \ast [r_2](Q)\}\;
  \texttt{mov}~\texttt{\%cr3},~r;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\end{mathpar}
Because the precondition on this rule is global, we avoid issues with framing. A more typical frame rule can be recovered from Hoare doubles in the typical way~\cite{Chlipala2013Bedrock} A more typical frame rule can be recovered from Hoare doubles in the typical way~\cite{Chlipala2013Bedrock}
with the added restriction that \lstinline|cr3| is restored to its pre-framing value before un-framing (in this case, the framed-out assertion is effectively
stashed under an other-space assertion for the initial address space, and pulled out as part of restoring).

@@

\inferrule[WriteToRegFromVirtMem]{

@@

\inferrule[WriteToVirtMemFromReg]{

@@

\inferrule[WriteToRegCtlFromRegModal]{

@@

Figure \ref{fig:wpdamd} includes two  rules for accessing memory at an address stored in a register $r_a$. 
Ultimately, any memory access needs to ensure the relevant address translation would succeed,
which can be ensured by what we informally call the page-table-walk points-to collection
($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}).
\TirNameStyle{WriteToRegFromVirtMem} and \TirNameStyle{WriteToVirtMemFromReg}
each use a virtual-pointsto assertion ($\textsf{vaddr} \mapsto_{\textsf{v}} \textsf{v}$),
and are nearly-standard (assembly) separation logic rules for memory accesses.
However, because we split the physical resources for the page table walk from the
virtual points-to itself (in order to permit the physical page tables to be freely modified
as long as they preserve virtual-to-physical translations), the rule requires $\mathcal{I}\textsf{ASpace}$
for the current address space to be carried through.
While doing the proof of soundness for these rules,
the token ($\sumwalkabs\vaddr\qfrac\paddr$) from the virtual points-to
is used to extract to extract page-table-traversal points-to collection
for the relevant address from that resource,
and then those  physical points-to assertions describing the page table walk
(from $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$) are used to prove
the operational semantics succeed and have the desired effect.

@@

\subsection{Soundness}
Our rules from Figure \ref{fig:wpdamd} are proven sound in Iris against an assembly-level hardware model
implementing a fragment of x86-64 including 64-bit address translation with 4-level page tables.

At the moment our proofs do rely on 13 small axioms of properties which should be provable, but
are challenging to discharge due to some representation choices in our model.\footnote{See \lstinline|srx/x64/machine/current_axioms.v|}

>>>>

 Definition L4_L1_PointsTo (maddr l4e l3e l2e l1e paddr :word 64) : vProp $\Sigma$ := $\lambda$ cr3val.
  $\ulcorner$ entry_present l4e $\land$ entry_present l3e $\land$ entry_present l2e $\land$ entry_present l1e$\urcorner$ $\ast$
  (l4M52 maddr cr3val) $\sim$ (l4off maddr cr3val) $\mapsto_{a}$ {q1}  l4e  $\ast$
  (l3M52 maddr l4e) $\sim$ (l3off maddr l4e)  $\mapsto_{a}$ {q2}  l3e $\ast$ 
  (l2M52 maddr l3e) $\sim$ (l2off maddr l3e) $\mapsto_{a}$ {q3}  l2e $\ast$
  (l1M52 maddr l2e) $\sim$ (l1off maddr l2e) $\mapsto_{a}$ {q4}  l1e $\ast$ $\ulcorner$ addr_L1(va,l1e) = paddr $\urcorner$.
\end{lstlisting}
\vspace{-1em}
\caption{A Strong Virtual Points-to Relation}
  \label{fig:strongvirtualpointsto}
\end{figure*}

\subsection{An Overly-Restrictive Definition for Virtual Memory Addressing}
\label{sec:overly-restrictive}
A natural definition for a virtual points-to that depends on the current address space is to
require that in order for a virtual address \textsf{va} to point to a value \textsf{v}, the assertion contains
partial ownership of the physical memory involved in the page table walk that would translate \textsf{va} to
its backing physical location --- with locations existentially quantified since a virtual points-to should not assert
\emph{which} locations are accessed in a page table walk.
Figure \ref{fig:strongvirtualpointsto} gives such a definition.
It asserts the existence of four page table entries, one at each translation level, and via \lstinline|L4_L1_PointsTo|
asserts from the current \lstinline|cr3| value, taking the offsets from \textsf{va} (as in Figure \ref{fig:pagetables})
to index each level, finding the existentially quantified entries in each level, 
ending with the L1 entry, whose physical page base is added to the page offset of \textsf{va} to obtain the physical address
 --- and there (\lstinline|addr_L1(va,l1e)|) in physical memory, is the value.
Most of the definition lives directly in \textsf{vProp}, using the separation logic structure lifted from Iris's \textsf{iProp}.

@@

  \begin{mathpar}
  \inferrule{ }{\mathsf{GhostMap}(\gamma,\theta) \ast \ghostmaptoken{\gamma}{\mathsf{pa}}{\mathsf{va}} \vs 
  \mathsf{GhostMap}(\gamma,\theta[\mathsf{pa}\mapsto\mathsf{va'}]) \ast \ghostmaptoken{\gamma}{\mathsf{pa}}{\mathsf{va}'} 
  }
  \and
  \inferrule{ }{\mathsf{GhostMap}(\gamma,\theta) \ast \ghostmaptoken{\gamma}{\mathsf{pa}}{\mathsf{va}} \wand \ulcorner \theta(\mathsf{pa})=\mathsf{Some}(\mathsf{va}) \urcorner }
  \end{mathpar}

  % Updating the virtual-to-physical (ghost) mapping for a virtual address $va$
  % requires full ownership of the token. 
  Then the \emph{virtual memory manager's invariant} ensures that for each $\ghostmaptoken{\gamma}{va}{pa}$ mapping
  in this map, there are \emph{physical} resources sufficient to ensure that address translation for $va$
  will resolve in the hardware to $pa$ via --- $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$.
  This allows modification of the page tables themselves as long as the changes do not invalidate overall virtual-to-physical
  translation.
  Consistency follows from replacing the page table walk with ownership of the ghost map token for the relevant ghost map
  entry, preventing the overall translation from being changed separately from the virtual points-to.
  \looseness=-1

@@

  \begin{figure*}
  \footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}(\ptablestore,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}(\ptablestore,m) \ast \\
 \qquad\qquad\qquad \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e l3e l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \textsf{ where } 
   \textsf{ASpace\_Lookup} (\ptablestore,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
    %\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}
    \ptableabswalk{\delta,\theta}

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant with a fixed global map of address-space names $m$}
  \label{fig:peraspaceinvariant}
  \end{figure*}

@@

% In Figure \ref{fig:addrswitch}, we see two different address-spaces rooted at R0 and R1 in the pre/post condition. To be put into words, while switching an address-space by executing \lstinline|mov_ctl|) updating the  \lstinline|cr3|, it is expected to mention resources of two different address-spaces in the specification. In the precondition,  \lstinline|cr3| is loaded with R0 and all the virtual-pointsto assertions $a_0 \mapsto_{\textsf{v}}p_0$ and $a_1 \mapsto_{\textsf{v}}p_1$ are in the current-view of the memory and valid in the ambient logic \textsf{vProp}. The other address-space with the root address R1 has its own resources ($b_0 \mapsto_{\textsf{v}}p_3$ and $b_1 \mapsto_{\textsf{v}}p_4$) which are invalid under the current address space rooted at R0. This pattern of specification requires resources being mentioned under the \textit{world} they are valid, i.e. a modal-framing within the specification. Once the current address-space becomes the one with root address R1, i.e.  \lstinline|cr3| is loaded with R1, the context-resources $b_0 \mapsto_{\textsf{v}}p_3$ and $b_1 \mapsto_{\textsf{v}}p_4$ are loaded into the current-view of the memory, and all the other resources that are valid under R0 are introduced into the modal context rooted at R0
\begin{figure}
\footnotesize
\begin{mathpar}
    \mbox{$[r](\mathsf{P}) \;: \mathsf{vProp} \; \Sigma \; \stackrel{\triangle}{=} \; \lambda \_,\; \mathsf{P}~\mathsf{r} \vspace{0.5em}$}
    \and
    \mbox{$\textsf{Fact P} \stackrel{\triangle}{=} \;\forall \; ,r \; r' \ldotp  \textsf{P r} \dashv\vdash \textsf{P r'}$}
    \and
    \inferrule*{ }{\textsf{Fact}\;[r](\textsf{P}) }
    \and
    \inferrule*{ }{\textsf{Fact}\;(\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv})}
    \and
    \inferrule{ }{\textsf{Fact}\;(\textsf{w} \mapsto_{\textsf{p}} \{q\} \textsf{ v} )}
    \and
    \inferrule{ }{(\textsf{P} \vdash \textsf{Q}) \vdash  ([r](\textsf{P}) \vdash  [r](\textsf{Q}))}
    \and
    \inferrule{ }{[r](\textsf{P} \ast \textsf{Q}) \dashv\vdash ([r](\textsf{P}) \ast [r](\textsf{Q})}\vspace{0.5em}
    \and
    \inferrule{ }{[r](\textsf{P}\land \textsf{Q}) \dashv\vdash [r](\textsf{P})\land [r](\textsf{Q})}
    \and
    \inferrule{ }{[r](\textsf{P}\lor \textsf{Q}) \dashv\vdash [r](\textsf{P})\lor [r](\textsf{Q})}
    \and
    \inferrule{ }{\textsf{Fact P} \vdash  [r](\textsf{P}) \dashv\vdash (\textsf{P})} %(\textsf{vProp }\Sigma))\\
\end{mathpar}
\vspace{-1em}
  \caption{Other-space Modality and Its Laws}
  \label{fig:modaldef}
  \end{figure}
Figure \ref{fig:modaldef} gives the definition of our modal operator for asserting truth of a modal
(address-space-contingent) assertion \emph{in another address space}, which we call
the \emph{other-space} modality. The definition itself is not
particularly surprising --- as our modal assertions are semantically predicates on a page table root (physical)
address, the assertion $[r](P)$ is a modal assertion which ignores the (implicit) current page table root,
and evaluates the truth of $P$ as if $r$ were the page table root. 
The novelty here is not in the details of the definition, but in recognizing that this is the right way to deal with
multiple address spaces, and working out how to support interaction of multiple address spaces (discussed in the next section).
% This simple definition is 
% what we need in order for assertions in one address space to talk about assertions in another.
\looseness=-1

@@

We can prove that this modality follows certain basic laws: that its truth is independent of the address
space in which it is considered,  it distributes over various logical connectives, and it follows the rule of
consequence.
We call \textsf{vProp} assertions whose truth is independent of the current address space
\textsf{Fact}s; these include other-space assertions, physical memory points-tos, and register assertions.
\textsf{Fact}s can move in and out of other-space modalities freely.

@@

\subsection{Subtleties of Changing Address Spaces Using Modalities}
% \subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
% \subsubsection{Issues with Using Hoare Triples in Address-Space Specification}
\label{sec:issues}
Operating systems on x86-64 change address spaces by updating \lstinline|cr3|.
Conceptually, we expect a proof rule for such an update to take move resources in and out of modalities for
the old and new address spaces. This suggests a rule like the following, which captures the intuition of the
address space change, but is subtly unsound under common assumptions.
% An important subtlety arises with supporting \lstinline|mov|s into \lstinline|cr3|. Consider the hypothetical rule
\begin{mathpar}
\inferrule*[left=Broken]{ }{
  \{P \ast \textsf{cr3}\mapsto_{r} \;r_1 \ast r \mapsto_{r} \;r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast \textsf{cr3} \mapsto_{r} r_2 \ast r \mapsto_{r} r_2 \ast Q\}
}
\end{mathpar}
This rule takes all assertions not under a modality ($P$), and moves them under a modality for the old address space
(which is where they were valid), while pulling assertions for the new, target address space out of a corresponding modality.
(For clarity, we also omit the address space invariants, which must also be present.)
% which is slightly simplified (by dropping per-address-space resources) for clarity in making the following point.
% This rule captures the intuitive change of address space in a Hoare triple.
Unfortunately, as stated, this interacts quite poorly with the traditional frame
rule.
If we abbreviate the pre- and post-conditions above as \textsf{Pre} and \textsf{Post}:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\textsf{Pre}\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{\textsf{Post}\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \textsf{Pre}\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast \textsf{Post}\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. 
The essential problem is that the frame rule is motivated by local reasoning about local updates, but
a switch of address space is a \emph{global} change that may invalidate information about virtual addresses.
Thus framing around \lstinline|cr3| updates is unsound, which has important impacts on our formal development.

@@

% This is then how we justify our actual rule for \lstinline|cr3| updates, which is \emph{almost}
This leads to our actual rule for \lstinline|cr3| updates, which is \emph{almost}
\begin{mathpar}
\inferrule*[left=ChangeAddressSpace]{
  \{[r_1](P) \ast cr3 \mapsto_{r} r_2 \ast r \mapsto_{r} r_2 \ast Q\}\;\overline{is}
}{
  \{P \ast cr3 \mapsto_{r} r_1 \ast r \mapsto_{r} r_2 \ast [r_2](Q)\}\;
  \texttt{mov}~\texttt{\%cr3},~r;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\end{mathpar}
(Again, the actual rule in Section \ref{sec:selected_rules} includes the address space invariants.)
Because the precondition on this rule is global, we avoid issues with framing. 
A more typical frame rule can be recovered from Hoare doubles in the typical way~\cite{Chlipala2013Bedrock} 
with the added restriction that \lstinline|cr3| is restored to its pre-framing value before un-framing (in this case, the framed-out assertion is effectively
stashed under an other-space assertion for the initial address space, and pulled out as part of restoring).

@@

\inferrule[\footnotesize WriteToRegFromVirtMem]{

@@

\inferrule[\footnotesize WriteToVirtMemFromReg]{

@@

\inferrule[\footnotesize WriteToRegCtlFromRegModal]{

@@

\paragraph{Accessing Virtual Addresses}
Figure \ref{fig:wpdamd} includes two  rules for accessing memory at an address stored in a register $r_a$. 
Ultimately, any memory access needs to ensure the relevant address translation would succeed,
which can be ensured by what we informally call the page-table-walk points-to collection
($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}).
\TirNameStyle{WriteToRegFromVirtMem} and \TirNameStyle{WriteToVirtMemFromReg}
each use a virtual-pointsto assertion ($\textsf{vaddr} \mapsto_{\textsf{v}} \textsf{v}$),
and are nearly-standard (assembly) separation logic rules for memory accesses.
However, because we split the physical resources for the page table walk from the
virtual points-to itself (in order to permit the physical page tables to be freely modified
as long as they preserve virtual-to-physical translations), the rule requires $\mathcal{I}\textsf{ASpace}$
for the current address space to be carried through.
The soundness proofs for these rules extract
% the token ($\sumwalkabs\vaddr\qfrac\paddr$) from the virtual points-to
the token ($\fracghostmaptoken{\delta}{\vaddr}{\paddr}{\qfrac}$) from the virtual points-to,
use that to extract the physical page-table-traversal points-to collection describing
the page table walk for the relevant address ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$)
from the invariant ($\mathcal{I}\mathsf{ASpace}$), prove that the page table walk succeeds and
memory or registers are updated appropriately, before re-packing the invariant and virtual points-to resources.

@@

\subsection{Soundness}
Our rules from Figure \ref{fig:wpdamd} are proven sound in Iris against an assembly-level hardware model
implementing a fragment of x86-64 including 64-bit address translation with 4-level page tables.
\todo[inline]{
Our rules for control transfers (\lstinline|jmp|, \lstinline|call|, and \lstinline|ret|) are currently
axiomatized (with completely standard specifications~\cite{ni2007contexts,Chlipala2013Bedrock}) because Iris's built-in machinery does not provide
convenient ways to discard the current continuation. We are working to adapt existing machinery~\cite{de2023type} to prove their soundness.
Our soundness proofs for all other instructions (including the instructions above, arithmetic, other heap-accessing operations, push, and pop)
are axiom-free.
}
% At the moment our proofs do rely on 13 small axioms of properties which should be provable, but
% are challenging to discharge due to some representation choices in our model.\footnote{See \lstinline|srx/x64/machine/current_axioms.v|}
